Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-12 paper/main.tex
started inputting changes to paper [WIP]

\title{Memory for television episodes preserves event content while introducing new across-event similarities}
% I think this title still needs work. In this version of the paper, we don’t really focus on the across-event similarities. Let's revisit this once the edits are complete - AH

% Constructing mnemonic events from meaningfully structured experiences
% Transforming meaningfully structured experiences into mnemonic events

@@

\begin{abstract}
Our experiences unfold over time defining unique \textit{trajectories} through the relevant representational spaces.  By casting our life events as temporally evolving trajectories, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how experiences are segmented into discrete events, and how the contents of event sequences evolve over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a television episode.  The content of participants' recountings of events from the original episode closely matched the original episode's content. Further, we introduce two novel metrics for assessing memory quality (precision and distinctiveness), both of which relate to participants' ability to recapitulate the experience.  Lastly, we identified a network of brain structures that are sensitive to the ``shapes'' of ongoing experiences, and an overlapping network that is sensitive (at the time of encoding) to how people later remembered those experiences in relation to other experiences.  In this way, modeling the content of richly structured experiences can reveal how (geometrically and conceptually) those experiences are segmented into events and integrated into our memories of other experiences.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from all others, and labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of precise recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the ``essence" of an experience, or whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

@@

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: the \textit{precision} with which a participant recounts each event and 2) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to participants' overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled" measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode, and how participants would later recount it.

\section*{Results}
To characterize the temporally dynamic contents of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) time segments spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and," ``or," ``but," etc.) across all features and scenes as the ``vocabulary" for the topic model.  We then concatenated the sets of words across all features contained in overlapping, sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various pairs of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active" in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event and across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each video event by averaging the topic vectors across timepoints each event spanned (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 video events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}
% I'm all for citing software packages, but where what's the distinction between packages we do cite (seaborn, brainiak, sklearn, hypertools, quail, umap, etc.) and don't cite (wordcloud, nilearn, fastdtw, etc.)?
% - Paxton
% I also flagged the seaborn reference for removal here. Def not critical to understanding the figure - Andy

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we parsed the time-varying content of the video, to obtain similar estimates for each participant's recall, we treated each overlapping  ``window" of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original video, despite different participants using widely varying language to describe the same event, and that those descriptions may not match the original annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic" space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

>>>>

\title{Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories}

@@

\begin{abstract}
The ways our experiences unfold over time may be characterized by unique \textit{trajectories} through geometric representational spaces.  In word embedding spaces, for example, each coordinate reflects a concept and nearby coordinates (in Euclidean distance) reflect semantically related concepts.  We propose a framework for projecting naturalistic experiences into word embedding spaces, such that the conceptual content of each moment of an experience, and how different moments of the experience relate, are reflected by the \textit{shape} of the experience's trajectory.  By projecting memories of those experiences into the same spaces, one may then geometrically compare the shape of the original experience's trajectory to the shape of how it is remembered later.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative questions about how we remember naturalistic experiences into quantitative geometric comparisons.  We applied our framework to data collected as participants watched, and later verbally recounted, a television episode while undergoing functional neuroimaging.  We found that all participants' remembered trajectories reflected the high-level \textit{essence} (i.e., large-scale narrative features) of the episode's trajectory, but participants differred markedly in their memories for low-level details (i.e, small-scale details of the trajectory).  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains transform our ongoing experiences when we encode them into episodic memories, and provides a formal geometric framework for characterizing the complex dynamic content of naturalistic experiences.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences and labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of precise recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the \textit{essence} of an experience, or whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

@@

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: (1) the \textit{precision} with which a participant recounts each event and (2) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled'' measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode, and how participants recount it.

\section*{Results}
To characterize the dynamic content of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are \textit{topics} (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) time segments spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active" in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each video event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 video events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}
% I'm all for citing software packages, but where what's the distinction between packages we do cite (seaborn, brainiak, sklearn, hypertools, quail, umap, etc.) and don't cite (wordcloud, nilearn, fastdtw, etc.)?
% - Paxton
% I also flagged the seaborn reference for removal here. Def not critical to understanding the figure - Andy

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we parsed the time-varying content of the video, to obtain similar estimates for each participant's recall, we treated each overlapping  ``window" of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original video, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the video annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

% JRM STOPPED HERE...
